\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-08-17 paper.tex
remove tilde g from main paper

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cG}{\mathcal G}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\cU}{\mathcal U}
\newcommand{\cR}{\mathcal R}
\newcommand{\cH}{\mathcal H}
\newcommand{\cB}{\mathcal B}
\newcommand{\cS}{\mathcal S}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}. \]

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, b_i^l) \Big) \\
  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l) \Big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g_i(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g_i$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g_i(b_i, G_i) = \tilde g_i(b_i)$ (for instance $\tilde g_i: b_i \mapsto b_i^\top D^{-1}b_i$), or $g_i(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{align}
  \label{eq:Lambda_ik-b_i|y_i}
  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} &= f(y_i | G_i = k) \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i \\
  \label{eq:Lambda_ik-b_i}
  &= \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i
\end{align}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-b_i|y_i} or~\eqref{eq:Lambda_ik-b_i} is not tractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S_{ik}^{(w)}$ used for the approximation.

@@

@@

The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
\begin{equation}
  \label{eq:def-hat-E-g-tilde}
  \hat \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde-1-2}
  \left\{
    \begin{array}{ll}
        \tilde g_{i, s, \gamma_k}^1&: b_i \mapsto \exp \big\{\tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
        \tilde g_{s, \gamma_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l).
    \end{array}
  \right. 
\end{equation}
We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.

Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{equation*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
\end{equation*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$. The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k).
\end{equation}
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. The closed-form update is then given by

\begin{equation}
  \label{eq:update_beta}
  {\beta_k^l}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} (y_i^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} U_{il} \Big]}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k}^1$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

>>>>

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cG}{\mathcal G}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\sA}{\mathscr A}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\cU}{\mathcal U}
\newcommand{\cR}{\mathcal R}
\newcommand{\cH}{\mathcal H}
\newcommand{\cB}{\mathcal B}
\newcommand{\cS}{\mathcal S}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\sA}, \] with $\sA = \sum_{a=1}^\cA \imath_a$ and so that
\begin{equation}
  \label{eq:def-asso-features}
  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
\end{equation}
where $\Psi(t) \in \R^{L\sA}$ denotes the association features vector at time $t$.

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g(b_i, G_i) = \tilde g(b_i)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{align}
  \label{eq:Lambda_ik-b_i|y_i}
  \Lambda_{ik,\tilde g}^{\theta^{(w)}} &= f(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i \\
  \label{eq:Lambda_ik-b_i}
  &= \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i
\end{align}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-b_i|y_i} or~\eqref{eq:Lambda_ik-b_i} is not tractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S_{ik}^{(w)}$ used for the approximation.

@@

\texttt{define $\Sigma_i$}

@@

The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
\begin{equation}
  \label{eq:def-hat-E-g-tilde}
  \hat \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.

Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{equation*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
\end{equation*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$. The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k).
\end{equation}
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{equation}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ U_{i}^\top \Sigma_i^{(w)} I_{n_{i}} (y_i - U_{i}\beta_k - V_{i}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big].
\end{equation}
The closed-form update is then given by

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\subsection{\texttt{QNMCEM class}}

Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto b_i g^4_{s, \gamma_k}(b_i), \\
    \tilde g^6_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation}
